---
title: "dcco_bcnh_model_markdown"
output: html_document
date: "2024-10-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r libraries, include=FALSE}
library(tidyverse)
library(R2jags)
library(bayesplot)

setwd("C:/Users/baill/OneDrive/r-projects/cormorant-colony-analysis/current working files/dcco_bcnh_thesis")
```


```{r load the data}
# Load the data
library(readr)
 model_data <- read_csv("C:/Users/baill/OneDrive/r-projects/cormorant-colony-analysis/current working files/dcco_bcnh_thesis/combined_dataset.csv", show_col_types = FALSE)
```

# Prepare the data
Select columns from combined dataset to use as fixed effects.

```{r data prep}
prepared_data <- model_data %>%
  select(year, bcnh_growthindex, total_nest_density, bcnh_nest_density, 
         dcco_growthindex, bcnh_nest_success, dcco_usurpation, 
         winter_nestremoval, deterrence_activenestremoval, deterrence_night,
         bcnh_road_proximity, raccoon_predation)

```

#impute missing data
```{r}
# Function to impute missing data based on variable type
impute_missing <- function(x, var_type) {
  if (all(!is.na(x))) return(x)
  
  observed <- x[!is.na(x)]
  n_missing <- sum(is.na(x))
  
  if (var_type == "normal") {
    mean_val <- mean(observed)
    sd_val <- sd(observed)
    x[is.na(x)] <- rnorm(n_missing, mean_val, sd_val)
  } else if (var_type == "exponential") {
    rate <- 1 / mean(observed)
    x[is.na(x)] <- rexp(n_missing, rate)
  } else if (var_type == "bernoulli") {
    prob <- mean(observed)
    x[is.na(x)] <- rbinom(n_missing, 1, prob)
  } else if (var_type == "uniform") {
    min_val <- min(observed)
    max_val <- max(observed)
    x[is.na(x)] <- runif(n_missing, min_val, max_val)
  }
  
  return(x)
}

# Impute missing data
prepared_data <- prepared_data %>%
  mutate(
    # Normal distribution for nest density and growth indices
    bcnh_growthindex = impute_missing(bcnh_growthindex, "normal"),
    dcco_growthindex = impute_missing(dcco_growthindex, "normal"),
    bcnh_nest_density = impute_missing(bcnh_nest_density, "normal"),
    
    # Exponential distribution for total nest density
    total_nest_density = impute_missing(total_nest_density, "exponential"),
    
    # Bernoulli distribution for management variables
    winter_nestremoval = impute_missing(winter_nestremoval, "bernoulli"),
    deterrence_activenestremoval = impute_missing(deterrence_activenestremoval, "bernoulli"),
    deterrence_night = impute_missing(deterrence_night, "bernoulli"),
    
    # For other variables, we'll use normal distribution as a default
    bcnh_nest_success = impute_missing(bcnh_nest_success, "normal"),
    dcco_usurpation = impute_missing(dcco_usurpation, "normal"),
    bcnh_road_proximity = impute_missing(bcnh_road_proximity, "normal"),
    raccoon_predation = impute_missing(raccoon_predation, "normal")
  )

```

#scale variables and arrange by year
Scale numerical covariates, excluding year, to have standard deviation of 1 and mean of 0.
Ensure data is in chronological order.
```{r}
# Scale variables and arrange by year
prepared_data <- prepared_data %>%
  mutate(across(-year, scale)) %>%
  arrange(year)

# Create lagged bcnh_growthindex
prepared_data <- prepared_data %>%
  mutate(bcnh_growthindex_lag = lag(bcnh_growthindex))
```

#Define number of observations (N) and unique years (N_years)
```{r}

# Define the number of observations (N) and unique years (N_years)
N <- nrow(prepared_data)
year <- as.numeric(factor(prepared_data$year))
N_years <- length(unique(year))
```

#impute NA in bcnh_growthindex_lag with mean of bcnh_growthindex
```{r}
# Improved data preparation
prepared_data <- prepared_data %>%
  mutate(bcnh_growthindex_lag = lag(bcnh_growthindex)) %>%
  mutate(bcnh_growthindex_lag = ifelse(is.na(bcnh_growthindex_lag), 
                                       mean(bcnh_growthindex, na.rm = TRUE), 
                                       bcnh_growthindex_lag))
```

# Prepare data for JAGS
```{r}
# Update jags_data list
jags_data <- list(
  N = nrow(prepared_data),
  N_years = length(unique(prepared_data$year)),
  bcnh_growth_index = as.vector(prepared_data$bcnh_growthindex),
  bcnh_growth_index_lag = as.vector(prepared_data$bcnh_growthindex_lag),
  total_nest_density = as.vector(prepared_data$total_nest_density),
  bcnh_nest_density = as.vector(prepared_data$bcnh_nest_density),
  dcco_growth_index = as.vector(prepared_data$dcco_growthindex),
  bcnh_nest_success = as.vector(prepared_data$bcnh_nest_success),
  winter_nestremoval = as.vector(prepared_data$winter_nestremoval),
  deterrence_activenestremoval = as.vector(prepared_data$deterrence_activenestremoval),
  deterrence_night = as.vector(prepared_data$deterrence_night),
  dcco_usurpation = as.vector(prepared_data$dcco_usurpation),
  bcnh_road_proximity = as.vector(prepared_data$bcnh_road_proximity),
  raccoon_predation = as.vector(prepared_data$raccoon_predation),
  year = as.numeric(factor(prepared_data$year))
)

# Print summary to check the imputation and scaling
summary(prepared_data)
```
# Specify JAGS model
BCNH growth index as response variable

Fixed effects: "total_nest_density", "bcnh_nest_density",                                                                                        "dcco_growthindex", "bcnh_nest_success", 
                                  "winter_nestremoval", "deterrence_activenestremoval", 
                                  "deterrence_night", "dcco_usurpation", 
                                  "bcnh_road_proximity", "raccoon_predation"
random effects: year to account for temporal autocorrelation

```{r model}

model_string <- "
model {
  for (i in 1:N) {
    # BCNH growth index as response variable (observed data)
    bcnh_growth_index[i] ~ dt(mu[i], tau, nu)
    
    # Mean of the response variable
    mu[i] <- beta[1] * total_nest_density[i] +
             beta[2] * bcnh_nest_density[i] +
             beta[3] * dcco_growth_index[i] +
             beta[4] * bcnh_nest_success[i] +
             beta[5] * winter_nestremoval[i] +
             beta[6] * deterrence_activenestremoval[i] +
             beta[7] * deterrence_night[i] +
             beta[8] * dcco_usurpation[i] +
             beta[9] * bcnh_road_proximity[i] +
             beta[10] * raccoon_predation[i] +
             beta[11] * pow(total_nest_density[i], 2) +  # Quadratic term
             beta[12] * total_nest_density[i] * dcco_growth_index[i] +  # Interaction term
             alpha_year[year[i]] +
             phi * bcnh_growth_index_lag[i]  # AR(1) term using lagged variable
    
    # Generate replicated data
    y_rep[i] ~ dt(mu[i], tau, nu)
  }
  
  # Priors for fixed effects
  for (j in 1:12) {
    beta[j] ~ dnorm(0, 0.001)
  }
  
  # Priors for variance parameters
  sigma ~ dunif(0, 100)
  sigma_year ~ dunif(0, 100)
  
  # Convert sigma to precision
  tau <- pow(sigma, -2)
  tau_year <- pow(sigma_year, -2)
  
  # Prior for degrees of freedom of t-distribution
  nu ~ dgamma(2, 0.1)
  
  # Prior for AR(1) coefficient
  phi ~ dunif(-1, 1)
  
  # Random year effect
  for (t in 1:N_years) {
    alpha_year[t] ~ dnorm(0, tau_year)
  }
}
"

```

#Initial values function and parameters
```{r inits}
inits <- function() {
  list(
    beta = rnorm(12, 0, 0.1),
    sigma = runif(1, 0, 1),
    sigma_year = runif(1, 0, 1),
    alpha_year = rnorm(N_years, 0, 0.1),
    nu = 5,
    phi = 0,
    z = rbinom(10, 1, 0.5),
    lambda = rexp(12, 1)
  )
}

parameters <- c("beta", "sigma", "sigma_year", "nu", "phi", "alpha_year", "y_rep")
```

#Fit the model
Starting with 50,000 iterations. Try 30,000 to 50,000.
```{r}
fit <- jags(
  data = jags_data,
  inits = inits,
  parameters.to.save = parameters,
  model.file = textConnection(model_string),
  n.chains = 3,
  n.iter = 50000,
  n.burnin = 2000,
  n.thin = 10
)
```

##diagnostics start
```{r}
library(coda)
library(parallel)
```

#convert to MCMC and posterior samples for bayesplot
```{r}
# Convert to MCMC for diagnostics
fit_mcmc <- as.mcmc(fit)

# Convert JAGS output to a format suitable for bayesplot
posterior_samples <- as.array(fit$BUGSoutput$sims.array)
```

#function to calculate Rhat and sum of squared residuals
```{r}
# Function to calculate Rhat
calculate_rhat <- function(mcmc_list) {
  gelman.diag(mcmc_list)$psrf[,1]
}

# Function to calculate sum of squared residuals
calc_ssr <- function(y, y_pred) {
  sum((y - y_pred)^2)
}
```

#calculate Rhat and print
Used to assess convergence of Markov chains
Ideal Rhat value: Generally, an Rhat value close to 1 indicates good convergence.
Common interpretations:
Rhat < 1.01 or 1.05: Excellent convergence
1.05 < Rhat < 1.1: Acceptable, but may warrant further investigation
Rhat > 1.1: Poor convergence, suggesting the chains have not mixed well
```{r}
# Calculate Rhat for all parameters
rhat_values <- calculate_rhat(fit_mcmc)

# Print Rhat values
print("Rhat values:")
print(rhat_values)

# Plot Rhat values
pdf("rhat_plot.pdf")
par(mar = c(8, 4, 2, 2))
barplot(rhat_values, las = 2, main = "Rhat Values", 
        ylab = "Rhat", cex.names = 0.7)
abline(h = 1.1, col = "red", lty = 2)
dev.off()
```

# Calculate P-values
```{r}
# Extract y_rep_samples
y_rep_samples <- fit$BUGSoutput$sims.list$y_rep

# Check the dimensions of  data
print(paste("Length of y_obs:", length(y_obs)))
print(paste("Dimensions of y_rep_samples:", paste(dim(y_rep_samples), collapse = " x ")))

# Function to calculate sum of squared residuals
calc_ssr <- function(y, y_pred) {
  sum((y - y_pred)^2)
}

# Calculate mean predictions
y_rep_mean <- apply(y_rep_samples, 2, mean)

# Calculate SSR for observed data
ssr_obs <- calc_ssr(y_obs, y_rep_mean)

# Calculate SSR for each replicated dataset
ssr_rep <- apply(y_rep_samples, 1, function(y) calc_ssr(y, y_rep_mean))

# Calculate Bayesian p-value
bayes_p_value <- mean(ssr_rep > ssr_obs)
print(paste("Bayesian P-value (using sum of squared residuals):", bayes_p_value))

# Plot histogram of replicated SSR with observed SSR
pdf("ssr_histogram.pdf")
hist(ssr_rep, main="Histogram of Replicated SSR", xlab="SSR")
abline(v=ssr_obs, col="red", lwd=2)
legend("topright", legend=c("Observed SSR"), col="red", lwd=2)
dev.off()

# Scatter plot of observed vs. predicted values
pdf("observed_vs_predicted.pdf")
plot(y_obs, y_rep_mean, 
     xlab = "Observed", ylab = "Predicted", 
     main = "Observed vs. Predicted Values")
abline(0, 1, col="red")
dev.off()

# Calculate and plot residuals
residuals <- y_obs - y_rep_mean
pdf("residual_plots.pdf")
par(mfrow=c(2,2))
plot(y_rep_mean, residuals, xlab="Fitted values", ylab="Residuals", main="Residuals vs Fitted")
abline(h=0, col="red")
qqnorm(residuals)
qqline(residuals, col="red")
hist(residuals, main="Histogram of Residuals")
plot(y_obs, residuals, xlab="Observed values", ylab="Residuals", main="Residuals vs Observed")
abline(h=0, col="red")
dev.off()

```

#trace plots and density plots
```{r}
# Load required library
library(bayesplot)

# Convert JAGS output to a format suitable for bayesplot
posterior_samples <- as.array(fit$BUGSoutput$sims.array)

# Create trace plots
pdf("trace_plots.pdf", width = 12, height = 8)
mcmc_trace(posterior_samples, 
           pars = c("beta[1]", "beta[2]", "beta[3]", "beta[4]", "beta[5]",
                    "beta[6]", "beta[7]", "beta[8]", "beta[9]", "beta[10]",
                    "beta[11]", "beta[12]", "sigma", "sigma_year", "nu", "phi"),
           facet_args = list(ncol = 4, strip.position = "top"))
dev.off()

#alpha_year parameters
pdf("trace_plots_with_alpha.pdf", width = 12, height = 12)
mcmc_trace(posterior_samples, 
           pars = c("beta[1]", "beta[2]", "beta[3]", "beta[4]", "beta[5]",
                    "beta[6]", "beta[7]", "beta[8]", "beta[9]", "beta[10]",
                    "beta[11]", "beta[12]", "sigma", "sigma_year", "nu", "phi",
                    "alpha_year[1]", "alpha_year[2]", "alpha_year[3]", "alpha_year[4]", "alpha_year[5]"),
           facet_args = list(ncol = 4, strip.position = "top"))
dev.off()

# To see all available parameters
print(dimnames(posterior_samples)[[3]])

# Create density plots
pdf("density_plots.pdf", width = 12, height = 8)
mcmc_dens(posterior_samples,
          pars = c("beta[1]", "beta[2]", "beta[3]", "beta[4]", "beta[5]",
                   "beta[6]", "beta[7]", "beta[8]", "beta[9]", "beta[10]",
                   "beta[11]", "beta[12]","sigma", "sigma_year", "nu", "phi"),
          facet_args = list(ncol = 4))
dev.off()

# Create a combined plot with both traces and densities
pdf("combined_diagnostic_plots.pdf", width = 15, height = 12)
mcmc_combo(posterior_samples,
           pars = c("beta[1]", "beta[2]", "beta[3]", "beta[4]", "beta[5]",
                    "beta[6]", "beta[7]", "beta[8]", "beta[9]", "beta[10]",
                    "beta[11]", "beta[12]", "sigma", "sigma_year", "nu", "phi"),
           combo = c("dens", "trace"),
           facet_args = list(ncol = 4))
dev.off()

```

#Plot residuals and save summary of posterior distributions
```{r}
# Plot residuals
pdf("residual_plots.pdf")
par(mfrow = c(2, 2))
hist(residuals, main = "Histogram of Residuals", xlab = "Residuals")
plot(y_obs, residuals, main = "Residuals vs Observed", 
     xlab = "Observed Values", ylab = "Residuals")
abline(h = 0, col = "red")
qqnorm(residuals)
qqline(residuals, col = "red")
plot(fitted(fit), residuals, main = "Residuals vs Fitted", 
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")
dev.off()

# Summary of posterior distributions
summary(fit_mcmc)

# Save the summary results as a CSV file
posterior_samples <- as.matrix(fit_mcmc)
results <- data.frame(
  parameter = colnames(posterior_samples),
  mean = colMeans(posterior_samples),
  lower_ci = apply(posterior_samples, 2, quantile, probs = 0.025),
  upper_ci = apply(posterior_samples, 2, quantile, probs = 0.975)
)

# Print and save results table
print(results)
write.csv(results, "jags_comprehensive_model_results_table.csv", row.names = FALSE)
```


#calculate,check, visualize residuals
```{r}
# Calculate residuals
y_obs <- jags_data$bcnh_growth_index
mu <- fit$BUGSoutput$sims.list$mu
mu_mean <- apply(mu, 2, mean)
residuals <- y_obs - mu_mean

# Check residuals
print(residuals)

# Visualize residuals
pdf("residual_plots.pdf")
hist(residuals, main="Histogram of Residuals", xlab="Residuals")
plot(y_obs, residuals, main="Residuals vs Observed", xlab="Observed Values", ylab="Residuals")
abline(h=0, col="red")
dev.off()
```


#5-fold cross validation
```{r}
# Perform 5-fold cross-validation
k <- 5
folds <- cut(seq(1, N), breaks = k, labels = FALSE)
cv_results <- vector("list", k)

for (i in 1:k) {
  test_indices <- which(folds == i)
  train_indices <- which(folds != i)
  
  # Prepare training data
  train_data <- jags_data
  train_data$N <- length(train_indices)
  train_data$bcnh_growth_index <- jags_data$bcnh_growth_index[train_indices]
  train_data$c <- jags_data$c[train_indices,]
  train_data$year <- jags_data$year[train_indices]
  
  # Fit model on training data
  cv_fit <- jags(data = train_data, inits = inits, parameters.to.save = parameters,
                 model.file = textConnection(model_string),
                 n.chains = 3, n.iter = 10000, n.burnin = 1000, n.thin = 10)
  
  # Predict on test data
  test_data <- jags_data
  test_data$N <- length(test_indices)
  test_data$bcnh_growth_index <- jags_data$bcnh_growth_index[test_indices]
  test_data$c <- jags_data$c[test_indices,]
  test_data$year <- jags_data$year[test_indices]
  
  pred <- jags.predict(cv_fit, test_data)
  
  # Calculate RMSE
  rmse <- sqrt(mean((jags_data$bcnh_growth_index[test_indices] - colMeans(pred))^2))
  cv_results[[i]] <- rmse
}

mean_cv_rmse <- mean(unlist(cv_results))
print(paste("Mean cross-validated RMSE:", mean_cv_rmse))
```
